Cartograms with Rivers

### 2022-11-08 paper/GHRVis.tex
draft

@@

\begin{abstract}
Cartograms are hybrid representations of geographical and abstract data based on a value-by-area mapping technique. Based on the Dorling cartogram, the Demers cartogram uses squares instead of circles to represent regions, enabling the comparison of regions in a more intuitive way with a higher screen space utilization. One of the disadvantages of the Dorling cartogram and its variants is that regions may be displaced from their original positions, reducing the legibility, readability, and geographical accuracy. To address this, we propose a novel cartogram layout algorithm that integrates dynamic topological features such as rivers into Demers cartograms. Rivers affect both the layout and appearance of cartograms. We evaluate the proposed layout algorithm through a user study with an application to an Electronic Health Records (EHR) dataset. The results demonstrate that the proposed layout algorithm is able to preserve some of the original properties of the cartogram, while improving the legibility, readability, and geographical accuracy of the cartogram. 
\end{abstract}

>>>>

\usepackage[nameinlink]{cleveref}

@@

\begin{abstract}
Cartograms are hybrid representations of geographical and abstract data based on a value-by-area mapping technique. A relative of the Dorling cartogram, the Demers cartogram uses squares instead of circles to represent regions, enabling the comparison of regions in a more intuitive way with a higher screen space utilization. One of the disadvantages of the Dorling cartogram and its variants is that regions may be displaced from their original positions, reducing the legibility, readability, and geographical accuracy. To address this, we propose a novel cartogram layout algorithm that integrates dynamic topological features such as rivers into Demers cartograms. Rivers affect both the layout and appearance of cartograms. We evaluate the proposed layout algorithm through a user study with an application to an Electronic Health Records (EHR) dataset. The results demonstrate that the proposed layout algorithm is able to preserve some of the original properties of the cartogram, while improving the legibility, readability, and geographical accuracy of the cartogram. 
\end{abstract}

### 2022-11-08 paper/section/appendix.tex
draft: autoref algorithm
draft

\section{Pre-processing Shapefiles}

Shapefiles from different sources are likely to be incompatible. In our case, the NHS CCG shapefile is incompatible with the river shapefiles. The major reason for the incompatibility is the coordinate reference system (CRS). The CRS of the CCG shapefile is EPSG:27700 (OSGB36 - British National Grid). The CRS of the river shapefiles is EPSG:4326 (WGS84 - World Geodetic System). Here we provide some pre-processing steps using QGIS (version: 3.26.0-Buenos Aires) \cite{qgisWelcome} to handle the incompatibility issue and reduce shapefile size to improve performance.

\subsection{Import Shapefiles into QGIS}
We first load all three river shapefiles into QGIS \autoref{fig:import_rivers}, followed by the CCG shapefile \autoref{fig:import_ccgs}.

@@

We then use QGIS to unify the CRS, and export both layers in the GeoJSON format. See \autoref{fig:export_rivers} and \autoref{fig:export_ccgs}. The unified layer is shown in \autoref{fig:unify_crs}.

@@

We then merge two layers into one layer, and export it in the TopoJSON format using Mapshaper \cite{blochMapshaper}. Mapshaper also supports the simplification of GeoJSON shapefiles. See \autoref{fig:export_topojson}.

@@

\autoref{tbale:pre-processing_result} shows the pre-processing result. The reduction in file size is significant and greatly reduces the initialization time of our implementation.

@@

>>>>

\section{Pre-processing Shapefiles}
\label{app:pre-processing}
Shapefiles from different sources are likely to be incompatible. In our case, the NHS CCG shapefile is incompatible with the river shapefiles. The major reason for the incompatibility is the coordinate reference system (CRS). The CRS of the CCG shapefile is EPSG:27700 (OSGB36 - British National Grid). The CRS of the river shapefiles is EPSG:4326 (WGS84 - World Geodetic System). Here we provide some pre-processing steps using QGIS (version: 3.26.0-Buenos Aires) \cite{qgisWelcome} to handle the incompatibility issue and reduce shapefile size to improve performance.

\subsection{Import Shapefiles into QGIS}
We first load all three river shapefiles into QGIS \Cref{fig:import_rivers}, followed by the CCG shapefile \Cref{fig:import_ccgs}.

@@

We then use QGIS to unify the CRS, and export both layers in the GeoJSON format. See \Cref{fig:export_rivers} and \Cref{fig:export_ccgs}. The unified layer is shown in \Cref{fig:unify_crs}.

@@

We then merge two layers into one layer, and export it in the TopoJSON format using Mapshaper \cite{blochMapshaper}. Mapshaper also supports the simplification of GeoJSON shapefiles. See \Cref{fig:export_topojson}.

@@

\Cref{table:pre-processing_result} shows the pre-processing result. The reduction in file size is significant and greatly reduces the initialization time of our implementation.

@@

\section{Procedure: TestIntersection}
% \begin{noindent}
    \begin{algorithm}[htb!]
        \caption{Procedure to test if a node's translation path, $ \nodeLineNV $ intersects a river.}\label{alg:check river intersection}
        \textbf{Input:} \\
        $ \nodeLineNV \gets $ the node's translation path \\
        $ \river \gets $ a river feature \\

        \textbf{Output:} \\
        Returns $ True $ if the node crosses a river. \\

        \textbf{Local variables:} \\
        $ \nodeBoundingBox, \riverEdgeBoundingBox \gets $ the bounding boxes for $ \nodeLineNV $ and $ \Edge $ \\
        $ \Edge \gets $ an edge of $ \river $ \\

        \begin{algorithmic}[1]
            \Procedure{TestIntersection}{$ \nodeLineNV $, $ \river $}

            \ForEach{$ \Edge \in \river $}
                \State $ \nodeBoundingBox \gets $ GetBoundingBox ($ \nodeLineNV $)
                \State $ \riverEdgeBoundingBox \gets $ GetBoundingBox ($ \Edge $)

                \If{$ \nodeBoundingBox ~intersect~ \riverEdgeBoundingBox = True $}
                    \State \Return{$ \nodeLineNV ~intersect~ \Edge $}
                    \EndIf
            \EndFor

            \State \Return{$ False $}
            \EndProcedure
        \end{algorithmic}
    \end{algorithm}
    %\end{noindent}

### 2022-11-08 paper/section/data.tex
draft

Obtaining the heterogenous data can be challenging, especially when an EHR dataset is involved \cite{wang2021EHRa}. The first step is to obtain both geospatial boundaries and EHR data that match one another. The second step is to pre-process the EHR data to remove empty and erroneous values. The final step is to transform the data into a format that is suitable for cartograms.

\bobgraph{Geospatial Data}

@@

\bobgraph{Clinical Commissioning Groups Shapefile:} Clinical Commissioning Groups (CCGs) are the primary administrative and geographic unit of the National Health Service (NHS) in the UK \cite{nhsNHS}. The number of CCGs changes over time due to the NHS re-organization. The most up-to-date shapefile is available from the Open Geography Portalx \cite{opengeographyportalxOpen}. We decided to use the CCG shapefile from 2020 at the time of writing, due to the absence of public EHR data published based on the latest CCG re-organizations that took place in 2021 and 2022.

\bobgraph{River Features:} We used OpenStreetMap \cite{openstreetmapRelation} as our data source to obtain shapefiles for River Thames, River Trent, and River Great Ouse in England. These rivers were chosen as they are well-known rivers and pass through regions with dense populations, and provide informative geographical and topological cues. Other rivers cloud be chose, including additional and smaller rivers remains future work.

@@

After acquiring the shapefiles, we used QGIS \cite{qgisWelcome} to manually adjust projections, and convert them into GeoJSON files. Finally, mapshaper \cite{blochMapshaper} is used to merge and convert the GeoJSON files into a TopoJSON \cite{TopoJSON} file. TopoJSON eliminates redundant coordinates in the data, improving the rendering speed of our implementation. See \autoref{table:pre-processing_result} on page \pageref{table:pre-processing_result} for the pre-processing result. We describe the above steps in more detail in \autoref{app:pre-processing}.

\bobgraph{EHR Data}

We obtained the Clinical Commissioning Group Outcomes Indicator Set (CCG OIS) from NHS Digital \cite{nhsdigitalClinical}. The OIS is a set of indicators that are used to measure the quality of care and the associated health outcomes in the NHS. Some datasets include:
\begin{itemize}
    \item Under 75 mortality: cardiovascular disease, respiratory disease, liver disease, and cancer
    \item Emergency hospital admission: stroke, alcohol-specific admission and readmission, coronary heart disease, re-admissions within 30 days of discharge, children with lower respiratory tract infections
\end{itemize}

>>>>

Obtaining the heterogenous data can be challenging, especially when an EHR dataset is involved \cite{wang2021EHRa}. The first step is to obtain both geospatial boundaries and EHR data. The second step is to pre-process the EHR data to remove empty and erroneous values. The final step is to transform the data into a format that is suitable for cartograms.

@@

\bobgraph{Clinical Commissioning Groups Shapefile:} Clinical Commissioning Groups (CCGs) are the primary administrative and geographic unit of the National Health Service (NHS) in the UK \cite{nhsNHS}. The number of CCGs changes over time due to NHS re-organization. The most up-to-date shapefile is available from the Open Geography Portalx \cite{opengeographyportalxOpen}. We decided to use the CCG shapefile from 2020 at the time of writing, due to the absence of public EHR data published based on the latest CCG re-organizations that took place in 2021 and 2022.

\bobgraph{River Features:} We used OpenStreetMap \cite{openstreetmapRelation} as our data source to obtain shapefiles for River Thames, River Trent, and River Great Ouse in England. These rivers were chosen as they are well-known rivers and pass through regions with dense populations, and provide informative geographical and topological cues. Including additional and smaller rivers remains future work.

@@

After acquiring the shapefiles, we used QGIS \cite{qgisWelcome} to manually adjust projections, and convert them into GeoJSON files. Finally, mapshaper \cite{blochMapshaper} is used to merge and convert the GeoJSON files into a TopoJSON \cite{TopoJSON} file. TopoJSON eliminates redundant coordinates in the data, improving the rendering speed of our implementation. See \Cref{table:pre-processing_result} on page \pageref{table:pre-processing_result} for the pre-processing result. We describe the above steps in more detail in \Cref{app:pre-processing}.

\bobgraph{EHR Data: }We obtained the Clinical Commissioning Group Outcomes Indicator Set (CCG OIS) from NHS Digital \cite{nhsdigitalClinical}. The OIS is a set of indicators that are used to measure the quality of care and the associated health outcomes in the NHS. Some datasets include:
\begin{itemize}
    \item Under 75 mortality: cardiovascular disease, respiratory disease, liver disease, and cancer
    \item Emergency hospital admission: stroke, alcohol-specific admission and readmission, coronary heart disease, re-admissions within 30 days of discharge, children with lower respiratory tract infections
\end{itemize}

### 2022-11-08 paper/section/evaluation.tex
draft

\section{Evaluation}

\subsection{Hypotheses Formulation}

We formulate the following hypotheses to evaluate the effectiveness of our approach:

\textbf{H1:} The introduction of dynamic topological features can effectively improve the recognizability of CCGs.

\textbf{H2:} \dots

\subsection{Participants}

We recruited X participants, [add demographics here].

\subsection{Datasets}

@@

\begin{itemize}
    \item Under 75 mortality from cardiovascular disease
    \item Emergency admissions for alcohol-related liver disease
\end{itemize}

Datasets are projected on a map that consists of 135 CGGs in England.

\subsection{Tasks}

We conducted tasks to evaluate the effectiveness of our approach.

\textbf{T1:} The participants were asked to familiarize themselves with the design, including the user options and the interaction methods.
\textbf{T2:} The participants

\subsection{Questionnaire}

We asked participants to complete a questionnaire to collect their feedback. The questionnaire consisted of the following questions:

\begin{itemize}
    \item \dots
    \item \dots
\end{itemize}

>>>>

\newcommand{\pCount}{14 }
\section{User-Centered Evaluation}

We conduct a preliminary user study to evaluate the effectiveness of our approach.

\subsection{User Study Tasks}

We formulate the following hypotheses to design our user study:

\textbf{H1:} The introduction of rivers can improve the legibility and recognizability of cartograms.

\textbf{H2:} The introduction of rivers can reduce the time needed to locate a CCG in a cartogram.

We test these hypotheses using location based tasks. Given a standard choropleth on the left side of the screen and a cartogram on the right side, we ask user study participants to locate the given region in the choropleth in the cartogram.

\subsection{User Study Variables}

We discuss the variables of our user study in this section.

\subsubsection{Independant Variables}

The primary independent variable is the presence of rivers in the cartograms, which directly impacts the final layout of the cartogram. We set this parameter to either render rivers or not.

% {
% \renewcommand{\arraystretch}{1.5}
% \begin{table}[!tb]
% 	\centering
% 		\begin{tabulary}{\columnwidth}{|*{2}{c|}}
% 			\hline
% 			\cellcolor{Mycolor2}\textbf{River Crossing} &
% 			\cellcolor{Mycolor2}\textbf{River Presence} \\
% 			\hline
% 			Allow & Present \\
% 			\hline
% 			Allow & Absent \\
% 			\hline
% 			Disallow & Present \\
% 			\hline
% 			Disallow & Absent \\
% 			\hline
% 		\end{tabulary}
% 	\caption{The combinations of independent variables.}
% 	\label{table:Independant Variables}
% \end{table}
% }

\subsubsection{Dependant Variables}

\bobgraph{Accuracy:} Given a CCG location in the choropleth on the left, we ask the participant to locate the corresponding node in the cartogram. The is the primary dependent variable measured by the number of correct CCGs chosen by the participants.

\bobgraph{Response time:} Another dependant variable is the time taken by the participants to complete each task.

\subsubsection{Control Variables}
\bobgraph{Choice of color map:} We use D3's built-in interpolateRdYlGn color map, a diverging color scheme of red, yellow, and green, to depict the data in our cartograms.

\bobgraph{Communicating the target CCG to the user:} We inform the participant about the target CCG that they need to find. The target CCG is shown in the form of a non-stop blinking (between its original color and black) area on the screen, each blink is given a duration of 2 seconds.

\subsection{User Study}

\subsubsection{Participants}

We recruited \pCount participants, ...

\subsubsection{Datasets}

@@

\begin{itemize}
    \item Population
    \item Under 75 mortality from cardiovascular disease
    \item Emergency admissions for alcohol-related liver disease
    \item Alcohol-specific admission and readmission
\end{itemize}

135 CGGs are rendered on the screen as a choropleth on the left. We then generated another view on the right using cartograms. The color in both visual designs is mapped to the dataset. See \Cref{fig:task}.

{
    \begin{figure}[htb!]
        \centering
        \includegraphics[width=\columnwidth,keepaspectratio]{figure/evaluation/task.png}
        \caption{A sample task for participants. The left shows the choropleth map, and the right shows the corresponding cartogram. Both images show the three longest rivers in England, and the color is mapped to CCG population. The target CGG is blinking on the choropleth (shown in black), participants are asked to identify this CCG on the cartogram.}
        \label{fig:task}
    \end{figure}
}

\subsubsection{Procedure}

The user study is designed to be conducted online, and participants are asked to complete the tasks using their own computer. The user study includes four parts:

\textbf{P1:} The participants are asked to listen to instructions and training provided as both text and videos. The instructions are designed to help participants understand the concepts used in the tasks. Instructional videos are available at \url{https://www.youtu.be/playlist?list=PLL7sHvxLtD75fMtrUQrAdddjt3wfFkcWz}.

\textbf{P2:} The participants are given three practice tasks to familiarize themselves with the user study design. A demonstration of three sample tasks is also included in the instructional video.

\textbf{P3:} The participants are asked to complete 16 location tasks. The response and reaction time are recorded. These 16 CCGs were carefully selected to avoid extreme cases (thus bias the result), in terms of size, color, and location.

\textbf{P4:} The participants were asked to complete a questionnaire that consists of Likert Scale questions and open-ended questions.

These are provided in the appendix, and available at \url{https://osf.io/q39w7}.

\subsection{User Study Results}

In this section we analyze the results of our user study. We recruited \pCount participants to perform 16 location tasks.

{
    \begin{figure}[htb!]
        \centering
        \includegraphics[width=\columnwidth,keepaspectratio]{figure/evaluation/task-rt.png}
        \caption{The scatter plot shows the response time (in seconds) and answers for each task, filtering out [add the exact number in the end] responses that took over 20 seconds.}
        \label{fig:task-rt}
    \end{figure}
}

### 2022-11-08 paper/section/introduction.tex
draft

{
    \begin{figure*}[ht!]
        \centering
        \includegraphics[width=0.8\textwidth,keepaspectratio]{figure/UI.png}
        \caption{A screenshot of the software. User options are provided to adjust the size, color mapping, visibility of nodes and rivers. Other options include the ability to control the overlap removal behavior: rivers can be static, dynamic, or constrained.}
        \label{fig:overview}
    \end{figure*}
}

@@

Among the four types of cartograms categorized by \citea{nusrat2016State} (contiguous, non-contiguous, rectangular, and Dorling), a trade-off between types of accuracy is made (See \autoref{table:accuracy}). For this project, we focus on non-contiguous cartograms like Demers cartograms, because they facilitate statistical comparison between regions and they can make good use of screen space. Building on Demers cartograms \cite{ian2002Cartogram}, we introduce and develop novel dynamic topological features, such as rivers, aiming to improve the readability and geographical accuracy without sacrificing statistical accuracy. Standard cartograms are composed of uniform shape nodes. As such, this can reduce their legibility. We implement a new cartographic layout algorithm that includes rivers into the layout of the nodes representing geographical regions. To reduce geographical errors and make efficient use of screen space, the algorithm also updates the position of rivers to accommodate the node layout. We then apply the algorithm to a real-world case study using an Electronic Health Records (EHR) dataset to evaluate the result. We also present a user study demonstrating its effectiveness.

@@

\begin{itemize}
    \item A new variant of Demers cartograms that incorporates dynamic topological features to improve readability and recognizability,
    \item A novel layout algorithm that preserves node positions relative to dynamic topological features such as rivers,
    \item A user study evaluation of the technique with an application to EHRs.
\end{itemize}

The results of the user study indicate that rivers can improve the legibility of cartograms. One of the major challenges involved is how to develop a layout algorithm that handles different shapes. In other words, the layout algorithm is novel because it handles different types of nodes -- rectangular representing regions and polylines representing rivers. Another challenge we overcome in developing the algorithm is to resolve stalemate situations introduced by dynamic topological features while minimizing geographical error.

>>>>

{
    \begin{figure*}[ht!]
        \centering
        \includegraphics[width=0.8\textwidth,keepaspectratio]{figure/UI.png}
        \caption{A screenshot of the software. User options are provided to adjust the size, color mapping, visibility of nodes and rivers. Other options include the ability to control the overlap removal behavior: rivers can be static, dynamic, or constrained [why too complicated? this is what we are doing in the software].}
        \label{fig:overview}
    \end{figure*}
}

@@

Among the four types of cartograms categorized by \citea{nusrat2016State} (contiguous, non-contiguous, rectangular, and Dorling), a trade-off between types of accuracy is made (See \Cref{table:accuracy}). For this project, we focus on non-contiguous cartograms like Demers cartograms, because they facilitate statistical comparison between regions and they can make good use of screen space. Building on Demers cartograms \cite{ian2002Cartogram}, we introduce and develop novel dynamic topological features, such as rivers, aiming to improve the readability and geographical accuracy without sacrificing statistical accuracy. Standard Demers cartograms are composed of uniform shape nodes. As such, this can reduce their legibility. We implement a new cartographic layout algorithm that includes rivers into the layout of the nodes representing geographical regions. Introducing rivers into the cartogram improves their legibility. To reduce geographical errors and make efficient use of screen space, the algorithm also updates the position of rivers to accommodate the node layout. We then apply the algorithm to a real-world case study using an Electronic Health Records (EHR) dataset to evaluate the result. We present a user study demonstrating its effectiveness.

@@

\begin{itemize}
    \item A new variant of Demers cartograms that incorporates rivers to improve readability and recognizability,
    \item A novel layout algorithm that preserves node positions relative to dynamic topological features such as rivers,
    \item A user study evaluation of the technique with an application to EHRs.
\end{itemize}

The results of the user study indicate that rivers can improve the legibility of cartograms. One of the major challenges involved is how to develop a layout algorithm that handles different shapes. In other words, the layout algorithm is novel because it handles different types of nodes -- rectangular representing regions and polylines representing rivers. Another challenge we overcome in developing the algorithm is to resolve stalemate situations introduced by dynamic rivers while minimizing geographical error.
